---
title: "harmonize-wq in R"
author: "Justin Bousquin, Cristina Mullin, Marc Weber"
date: '2022-08-31'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{harmonize-wq in R}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# Set chunk options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<br>

## Overview

Standardize, clean, and wrangle Water Quality Portal data into more analytic-ready formats using the harmonize_wq package. US EPA’s Water Quality Portal (WQP) aggregates water quality, biological, and physical data provided by many organizations and has become an essential resource with tools to query and retrieval data using python or R. Given the variety of data and variety of data originators, using the data in analysis often requires data cleaning to ensure it meets the required quality standards and data wrangling to get it in a more analytic-ready format. Recognizing the definition of analysis-ready varies depending on the analysis, the harmonize_wq package is intended to be a flexible water quality specific framework to help:

* Identify differences in data units (including speciation and basis)
* Identify differences in sampling or analytic methods
* Resolve data errors using transparent assumptions
* Reduce data to the columns that are most commonly needed
* Transform data from long to wide format

Domain experts must decide what data meets their quality standards for data comparability and any thresholds for acceptance or rejection.

<br>

<br>

## Installation & Setup

#### Install the harmonize-wq package (Command Line)

To install and set up the harmonize-wq package using the command line:

1. If needed, re-install [miniforge](https://github.com/conda-forge/miniforge). Once miniforge is installed. Go to your start menu and open the Miniforge Prompt.
2. At the Miniforge Prompt:
    - conda create --name wq_harmonize
    - activate wq_harmonize
    - conda install geopandas pip dataretrieval pint
    - may need to update conda
      - conda update -n base -c conda-forge conda
    - pip install harmonize-wq
    - pip install git+https://github.com/USEPA/harmonize-wq.git (dev version)

<br>

#### Install the harmonize-wq package (R)

**Alternatively**, you may be able to set up your environment and import the required Python packages using the block of R code below:

```{r, results = 'hide', eval=FALSE}
# If needed, install the reticulate package to use Python in R
install.packages("reticulate")
library(reticulate)

# The reticulate package will automatically look for an installation of Conda
# However, you may specify the location if needed using options(reticulate.conda_binary = 'dir')
options(reticulate.conda_binary = '~/AppData/Local/miniforge3/Scripts/conda.exe')

# Create a new Python environment called "wq-reticulate"
# Note that the environment name may need to include the full path (e.g. "~/AppData/Local/miniforge3/envs/wq_harmonize")
conda_create("wq-reticulate")

# Install the following packages to the newly created environment
conda_install("wq-reticulate", "geopandas")
conda_install("wq-reticulate", "pint")
conda_install("wq-reticulate", "dataretrieval")

# Install the harmonize-wq package
# This only works with py_install() (pip), which defaults to virtualenvs
# Note that the environment name may need to include the full path (e.g. "~/AppData/Local/miniforge3/envs/wq_harmonize")
py_install("harmonize-wq", pip = TRUE, envname = "wq-reticulate")

# To install the dev version of harmonize-wq from GitHub
# Note that the environment name may need to include the full path (e.g. "~/AppData/Local/miniforge3/envs/wq_harmonize")
py_install("git+https://github.com/USEPA/harmonize-wq.git@new_release_0-3-8", pip = TRUE, envname = "wq-reticulate")

# Specify the Python environment to be used
use_condaenv("wq_harmonize")

# Test that your Python environment is correctly set up
# Both imports should return "Module(package_name)"
import("harmonize_wq")
import("dataretrieval")
```

<br>

#### Import required libraries

The full list of dependencies that should be installed to use the harmonize-wq package can be found in [`requirements.txt`](https://github.com/USEPA/harmonize-wq/blob/new_release_0-3-8/requirements.txt). **Note that `reticulate::repl_python()` must be called to execute these commands using the reticulate package in R.**

```{r}
# Use reticulate to execute python commands
reticulate::repl_python()
```

```{python}
# Use these reticulate imports to test the modules are installed
import harmonize_wq
import dataretrieval
import os
import pandas
import geopandas
import dataretrieval.wqp as wqp
from harmonize_wq import harmonize
from harmonize_wq import convert
from harmonize_wq import wrangle
from harmonize_wq import clean
from harmonize_wq import location
from harmonize_wq import visualize
```

<br>

<br>

## Usage

The following example illustrates a typical harmonization process using the harmonize-wq package on WQP data retrieved from Perdido and Pensacola Bays, FL.

First, determine an area of interest (AOI), build a query, and retrieve water temperature and Secchi disk depth data from WQP for the AOI using the dataretrieval package:

```{python, message=FALSE, warning=FALSE, error=FALSE}
# File for area of interest (Pensacola and Perdido Bays, FL)
aoi_url = r'https://raw.githubusercontent.com/USEPA/harmonize-wq/main/harmonize_wq/tests/data/PPBays_NCCA.geojson'

# Build query and get WQP data with dataretrieval
query = {'characteristicName': ['Temperature, water',
                                'Depth, Secchi disk depth',
                                ]}

# Use harmonize-wq to wrangle
query['bBox'] = wrangle.get_bounding_box(aoi_url)
query['dataProfile'] = 'narrowResult'

# Run query
res_narrow, md_narrow = wqp.get_results(**query)

# DataFrane of downloaded results
res_narrow
```

Next, harmonize and clean all results:

```{python, message=FALSE, warning=FALSE, error=FALSE}
df_harmonized = harmonize.harmonize_all(res_narrow, errors='raise')
df_harmonized

# Clean up the datetime and sample depth columns
df_cleaned = clean.datetime(df_harmonized)
df_cleaned = clean.harmonize_depth(df_cleaned)
df_cleaned
```

There are many columns in the data frame that are characteristic specific, that is they have different values for the same sample depending on the characteristic. To ensure one result for each sample after the transformation of the data, these columns must either be split, generating a new column for each characteristic with values, or moved out from the table if not being used.

```{python, message=FALSE, warning=FALSE, error=FALSE}
# Split the QA_flag column into multiple characteristic specific QA columns
df_full = wrangle.split_col(df_cleaned)

# Divide table into columns of interest (main_df) and characteristic specific metadata (chars_df)
main_df, chars_df = wrangle.split_table(df_full)

# Combine rows with the same sample organization, activity, location, and datetime
df_wide = wrangle.collapse_results(main_df)

# Reduced columns
df_wide.columns
df_wide.head()
```

Finally, the cleaned and wrangled data may be visualized as a map:

```{python, message=FALSE, warning=FALSE, error=FALSE}
# Get harmonized stations clipped to the AOI
stations_gdf, stations, site_md = location.get_harmonized_stations(query, aoi=aoi_url)

# Map average temperature results at each station
gdf_temperature = visualize.map_measure(df_wide, stations_gdf, 'Temperature')
gdf_temperature.plot(column='mean', cmap='OrRd', legend=True)
```

<br>

<br>