---
title: "harmonize-wq in R"
format: html
editor: visual
author: "Justin Bousquin, Cristina Mullin, Marc Weber"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 8
    fig_width: 8
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{harmonize-wq in R}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Overview

Standardize, clean, and wrangle Water Quality Portal data into more
analytic-ready formats using the harmonize_wq package. US EPA's Water
Quality Portal (WQP) aggregates water quality, biological, and physical
data provided by many organizations and has become an essential resource
with tools to query and retrieval data using python or R. Given the
variety of data and variety of data originators, using the data in
analysis often requires data cleaning to ensure it meets the required
quality standards and data wrangling to get it in a more analytic-ready
format. Recognizing the definition of analysis-ready varies depending on
the analysis, the harmonize_wq package is intended to be a flexible
water quality specific framework to help:

-   Identify differences in data units (including speciation and basis)
-   Identify differences in sampling or analytic methods
-   Resolve data errors using transparent assumptions
-   Reduce data to the columns that are most commonly needed
-   Transform data from long to wide format

Domain experts must decide what data meets their quality standards for
data comparability and any thresholds for acceptance or rejection.

## Installation & Setup

Using R/reticulate requires an installation of Python to bind to, and EPA's harmonize-wq package must be run from a python environment with the packages it depends on.

For environment management, reticulate requires either conda (recommended) or virtualenv.

There are multiple installers available for [Conda](https://conda.io/projects/conda/en/latest/user-guide/install/index.html).
The examples use miniforge3, one of several verions of [miniforge](https://github.com/conda-forge/miniforge). In R, [miniconda](https://docs.anaconda.com/miniconda/) can be installed using reticulate::install_miniconda().

#### Option 1: Install the harmonize-wq Package Using the Command Line

To install and set up the harmonize-wq package using the command line:

1.  If needed, re-install
    [miniforge](https://github.com/conda-forge/miniforge). Once
    miniforge is installed. Go to your start menu and open the Miniforge
    Prompt.
2.  At the Miniforge Prompt, run:
    -   to update conda:

        > conda update -n base -c conda-forge conda

    -   conda create --name wq_harmonize

    -   conda activate wq_harmonize

    -   conda install dependencies (from requirements.txt):

        > conda install "numpy<2.0" "pandas<2.0" "geopandas>=0.10.2, <0.13" "pint>=0.18" "dataretrieval>=1.0, <1.0.5" "pip"

    -   pip install harmonize-wq (dev-version shown):

        > pip install git+<https://github.com/USEPA/harmonize-wq.git>

#### Option 2: Install the harmonize-wq Package Using R

**Alternatively**, you may be able to set up your environment and import
the required Python packages using R.

First, run the chunk below to install the reticulate package to use
Python in R.

```{r, results = 'hide'}
install.packages("reticulate", repos = "http://cran.us.r-project.org")
library(reticulate)
```

Once miniforge3 (or another Conda installer of your choice) is
installed, the reticulate package will automatically look for the
installation of Conda (conda.exe) on your computer.

```{r, results = 'hide'}
# options(reticulate.conda_binary = 'dir')
```

However, you may still need to specify the conda.exe location. To do so,
update the last line of code below to specify your conda.exe location

```{r, results = 'hide'}
# update the 'dir' in this chuck to specify the location of conda.exe on your computer
# Note: that the environment name may need to include the full path (e.g. "C:/Users/<USERNAME>/AppData/Local/miniforge3/Scripts/conda.exe")
# options(reticulate.conda_binary = "~/AppData/Local/miniforge3/Scripts/conda.exe")
```

Next, update the code chunk below to create a new Python environment in
the envs folder called "wq_harmonize". Note that the environment name
may need to include the full path (e.g.
"C:/Users/<USERNAME>/AppData/Local/miniforge3/envs/wq_harmonize")

```{r, results = 'hide'}
# reticulate::conda_create("~/AppData/Local/miniforge3/envs/wq_harmonize")
```

Install the following python packages to the newly created Python
environment called "wq_harmonize".

```{r, results = 'hide'}
# packages = c(
#              "numpy<2.0", "pandas<2.0", "geopandas>=0.10.2, <0.13", "pint>=0.18",
#              "dataretrieval>=1.0, <1.0.5", "pip"
#              )
# reticulate::conda_install("wq_harmonize", packages)
```

Uncomment to install EPA's harmonize-wq package most recent release or
development version.

```{r, results = 'hide'}
# Install the most recent release of the harmonize-wq package
# This only works with py_install() (pip = TRUE), which defaults to use virtualenvs
# reticulate::py_install("harmonize-wq", pip = TRUE, envname = "wq_harmonize")

# Install the development version of harmonize-wq from GitHub (optional)
#py_install("git+https://github.com/USEPA/harmonize-wq.git", pip = TRUE, envname = "wq_harmonize")
```

Specify the Python environment to be used, "wq_harmonize", and test that
your Python environment is set up correctly.

```{r}
# Specify environment to be used
#reticulate::use_condaenv("wq_harmonize")

# Test set up is correct
# Both imports should return "Module(package_name)"
reticulate::import("harmonize_wq")
reticulate::import("dataretrieval")
```

#### Import additional required libraries

The full list of dependencies that should be installed to use the
harmonize-wq package can be found in
[`requirements.txt`](https://github.com/USEPA/harmonize-wq/blob/new_release_0-3-8/requirements.txt).

```{python, results = 'hide'}
# Use these reticulate imports to test the modules are installed
import harmonize_wq
import dataretrieval
import os
import pandas
import geopandas
import dataretrieval.wqp as wqp
import pint
import mapclassify
import matplotlib.pyplot as plt
from harmonize_wq import harmonize
from harmonize_wq import convert
from harmonize_wq import wrangle
from harmonize_wq import clean
from harmonize_wq import location
from harmonize_wq import visualize
```

## harmonize-wq Usage: FL Bays Example

The following example illustrates a typical harmonization process using
the harmonize-wq package on WQP data retrieved from Perdido and
Pensacola Bays, FL.

**Note that `reticulate::repl_python()` must be called first to execute
these commands using the reticulate package in R.**

```{r, results = 'hide'}
# Use reticulate to execute python commands
reticulate::repl_python()
```

First, determine an area of interest (AOI), build a query, and retrieve
water temperature and Secchi disk depth data from the Water Quality
Portal (WQP) for the AOI using the dataretrieval package:

```{python, error = F}
# File for area of interest (Pensacola and Perdido Bays, FL)
aoi_url = r'https://raw.githubusercontent.com/USEPA/harmonize-wq/main/harmonize_wq/tests/data/PPBays_NCCA.geojson'

# Build query and get WQP data with dataretrieval
query = {'characteristicName': ['Temperature, water', 'Depth, Secchi disk depth',]}

# Use harmonize-wq to wrangle
query['bBox'] = wrangle.get_bounding_box(aoi_url)
query['dataProfile'] = 'narrowResult'

# Run query
res_narrow, md_narrow = wqp.get_results(**query)

# DataFrane of downloaded results
res_narrow
```

Next, harmonize and clean all results using the harmonize.harmonize_all,
clean.datetime, and clean.harmonize_depth functions.

Enter a ? followed by the function name, for example
?harmonize.harmonize_all, into the console for more details.

```{python, error = F}
df_harmonized = harmonize.harmonize_all(res_narrow, errors = 'raise')
df_harmonized

# Clean up the datetime and sample depth columns
df_cleaned = clean.datetime(df_harmonized)
df_cleaned = clean.harmonize_depth(df_cleaned)
df_cleaned
```

There are many columns in the data frame that are characteristic
specific, that is they have different values for the same sample
depending on the characteristic. To ensure one result for each sample
after the transformation of the data, these columns must either be
split, generating a new column for each characteristic with values, or
removed from the table if not needed.

```{python, error = F}
# Split the QA_flag column into multiple characteristic specific QA columns
df_full = wrangle.split_col(df_cleaned)

# Divide table into columns of interest (main_df) and characteristic specific metadata (chars_df)
main_df, chars_df = wrangle.split_table(df_full)

# Combine rows with the same sample organization, activity, location, and datetime
df_wide = wrangle.collapse_results(main_df)

# Reduced columns
df_wide.columns
df_wide.head()
```

Finally, the cleaned and wrangled data may be visualized as a map:

```{python, error = F}
# Get harmonized stations clipped to the AOI
stations_gdf, stations, site_md = location.get_harmonized_stations(query, aoi=aoi_url)

# Map average temperature results at each station
gdf_temperature = visualize.map_measure(df_wide, stations_gdf, 'Temperature')
plt.figure()
gdf_temperature.plot(column = 'mean', cmap = 'OrRd', legend = True)
plt.show()
```
